Differentiable Optics with {\Large $\partial$}Lux: I - Deep Calibration of Flat Field and Phase Retrieval with Automatic Differentiation

### 2022-12-16 src/tex/ms.tex
[showyourwork] overleaf sync

% Define document class
% \documentclass[twocolumn]{spie}
\documentclass[]{spieman}
% \documentclass[]{spie}
% \documentclass[twocolumn]{aastex631}
\usepackage[utf8]{inputenc}
\usepackage{showyourwork}
\usepackage{xspace}
%\documentclass[a4paper]{spie}  %>>> use this instead for A4 paper
%\documentclass[nocompress]{spie}  %>>> to avoid compression of citations

@@

\author[a,*]{Louis Desdoigts}
\author[b,c]{Benjamin J. S. Pope}
\author[a]{Peter G. Tuthill}
\author[b]{Jordan Dennis}

@@

In this series of papers, we present a new software package, \dlux, designed for fitting high-dimensional parametrized physical optics models to astronomical data. By using the Python library \textsc{Jax}\cite{jax}, we obtain \ac{gpu} hardware acceleration, as well as automatic differentiation or `autodiff' \cite{Margossian2018} features that enable high-dimensional optimization and inference with gradient descent or \ac{hmc}. In the present paper, we focus on the particular problem of estimating the \ac{prf} and \ac{psf} simultaneously - i.e. joint flat field calibration and phase retrieval. In Paper II, we will show how autodiff enables improvements to hardware \textit{design}, directly calculating and optimizing the Fisher information and therefore fundamental figures of merit of an optical system end-to-end. We make \dlux available as open-source software on GitHub\footnote{\href{https://github.com/LouisDesdoigts/dLux}{github.com/LouisDesdoigts/dLux}}, and encourage interested readers to use and contribute to this package as a community resource. This paper has been compiled using ShowYourWork \cite{Luger2021}, so that all figures link to the code used to produce them.

@@

% Phase Retrieval - G-S, Fienup, review papers, kernel phase
Phase retrieval is the problem of inferring these aberrations from data\cite{schechtman2014}, which is in general ill-posed \cite{barnett2020}: because of the Hermitian symmetry of the Fourier transform, and because we measure \textit{intensity} and not electric field in optical astronomy, there is a large space of aberrations that would generate the same intensity \ac{psf}. Fortunately, this space can be restricted to physically-realistic solutions and readily solved by algorithms such as the Gerchberg-Saxton algorithm \cite{gerchberg1972}, using ideas from compressed sensing \cite{candes2011}, or by machine learning \cite{metzler2018,isil2019,nishizaki2020}.  Phase retrieval was memorably performed to infer and correct the serious aberration on the \textit{Hubble Space Telescope} mirror at launch \cite{hubble_phase_ret}. 
Earlier work in the vein of this paper has shown that the phase retrieval problem can be efficiently solved by taking a forwards model of physical optics, obtaining partial derivatives with autodiff, and optimizing its parameters by gradient descent \cite{jurling_fienup,phase_ret_and_design}.

@@

In this paper we introduce a new optical simulation package, \dlux, available under an open-source BSD 3 license. It significantly extends the capabilities of \texttt{morphine} \cite{pope2021,phase_ret_and_design}, a previous \jax optics package from our team based on \texttt{poppy}, by taking full advantage of more-recent \jax features and libraries.  It is built in an object-oriented \jax framework called \equinox \cite{kidger2021equinox}, with an optical system consisting of an \equinox module wrapping a stack of `layers' which are themselves \equinox modules applying phase or amplitude screens, or performing Fourier or Fresnel transforms on a wavefront. We use the optimization library \optax \cite{optax2020github} for gradient-descent optimization, and the probabilistic programming library \numpyro \cite{Phan2019} for \ac{hmc}.

% Autodiff Optics
There are several other frameworks for automatically-differentiable optics simulation:
% WaveBlocks
the most similar is WaveBlocks \cite{page2020}, a PyTorch package for object-oriented modelling of fluorescence microscopy. 
% Sitzmann etc
The DeepOptics package \cite{sitzmann2018} has been developed in TensorFlow for camera design and computational imaging, and
% Liaudat
WaveDiff in TensorFlow has been used for detector calibration \cite{liaudat2021} and \ac{psf} modelling \cite{Liaudat2022}. 
We believe that \dlux fills a unique role in this ecosystem as the only fully-generalised modelling framework for physical optics. This has driven a design focus on flexibility and ease of open-source contribution and development. The goal is to provide a general object-oriented framework that enables all those working within optics to harness the advantages of autodiff and hardware acceleration provided by \jax with its user-friendly Numpy-like API.

@@

The optimisation is performed for a total of 100 epochs, with each gradient calculation taking $\sim$ 2.75 seconds to evaluate on an Apple M1 Max CPU, leading to a total optimisation time of $\sim$ 4.5 minutes.

This optimisation strategy performs well, with all of the parameters being recovered down to the noise level. This can be seen in Figure~\ref{fig:data_residual} with the right panel showing a sample residual of the data and model after optimisation, which is clearly only noise. Correlations and residual plots of the positions and fluxes after the optimisation are shown in Figure~\ref{fig:astro_params}. The 1-$sigma$ error bars 
are generated by harnessing the differentiablility of the model by calculating the covaraince matrix under the Laplace approximation, defined as the negative matrix inverse of the hessian of the likelihood function evaluated at the maximum likelihood estimate, ie the final recovered values. These indicate a poor recovery, however they are only marginalised over positions, fluxes and aberration, not the \ac{prf} which is the dominant noise source as its high dimensionality makes marginalisation computationally intractable. These parameters are recovered to the expected distribution when performed without modelling the \ac{prf}. Figure~\ref{fig:aberrations} shows the recovery of the optical aberrations, with a final residual aberration of 
% \textcolor{red}{showyourwork variable}
\variable{output/rms_opd_resid.txt} nm RMS. 
Finally, Figure~\ref{fig:flat_field} shows a correlation plot of the true and recovered sensitivity values and the corresponding residual histogram. The points of the correlation plot are colour-coded with the total counts for each pixel, showing as expected that pixels with higher photon counts having a better recovery of the PRF since these pixels have a greater \ac{snr}. Overall the parameters are very well recovered down to the noise level.

@@

\section{Discussion}

>>>>

% Define document class
% \documentclass[twocolumn]{spie}
\documentclass[]{spieman}
% \documentclass[]{spie}
% \documentclass[twocolumn]{aastex631}
\usepackage[utf8]{inputenc}
\usepackage{showyourwork}
\usepackage{xspace}
\usepackage{lineno}
\linenumbers
%\documentclass[a4paper]{spie}  %>>> use this instead for A4 paper
%\documentclass[nocompress]{spie}  %>>> to avoid compression of citations

@@

\author[a,*]{Louis Desdoigts}
\author[b,c]{Benjamin J. S. Pope}
\author[b]{Jordan Dennis}
\author[a]{Peter G. Tuthill}

@@

In this series of papers, we present a new software package, \dlux, designed for fitting high-dimensional parametrized physical optics models to astronomical data. By using the Python library \textsc{Jax}\cite{jax}, we obtain \ac{gpu} hardware acceleration, as well as automatic differentiation or `autodiff' \cite{Margossian2018} features that enable high-dimensional optimization and inference with gradient descent or \ac{hmc}. In the present paper, we focus on the particular problem of estimating the \ac{prf} and \ac{psf} simultaneously - i.e. joint flat field calibration and phase retrieval. In Paper~II, we will show how autodiff enables improvements to hardware \textit{design}, directly calculating and optimizing the Fisher information and therefore fundamental figures of merit of an optical system end-to-end. And in Paper~III, we will use this for end-to-end deconvolution of high angular resolution imaging with an unknown \ac{psf}. We make \dlux available as open-source software on GitHub\footnote{\href{https://github.com/LouisDesdoigts/dLux}{github.com/LouisDesdoigts/dLux}}, and encourage interested readers to use and contribute to this package as a community resource. This paper has been compiled using ShowYourWork \cite{Luger2021}, so that all figures link to the code used to produce them.

@@

% Phase Retrieval - G-S, Fienup, review papers, kernel phase
Phase retrieval is the problem of inferring these aberrations from data\cite{schechtman2014}, which is in general ill-posed \cite{barnett2020}: because of the Hermitian symmetry of the Fourier transform, and because we measure \textit{intensity} and not electric field in optical astronomy, there is a large space of aberrations that would generate the same intensity \ac{psf}. Fortunately, this space can be restricted to physically-realistic solutions and readily solved by algorithms such as the Gerchberg-Saxton algorithm \cite{gerchberg1972}, using ideas from compressed sensing \cite{candes2011}, or by machine learning \cite{metzler2018,isil2019,nishizaki2020}.  Phase retrieval was memorably performed to infer and correct the serious aberration on the \textit{Hubble Space Telescope} mirror at launch \cite{hubble_phase_ret}. 
Earlier work in the vein of this paper has shown that the phase retrieval problem can be efficiently solved even in the case of detector nonlinearity by taking a forwards model of physical optics, obtaining partial derivatives with autodiff, and optimizing its parameters by gradient descent \cite{jurling_fienup,phase_ret_and_design}, including a conference presentation of an early version of the present work \cite{Desdoigts2022}.

@@

In this paper we introduce a new optical simulation package, \dlux, available under an open-source BSD 3 license. It significantly extends the capabilities of \texttt{morphine} \cite{pope2021,phase_ret_and_design}, a previous \jax optics package from our team based on \texttt{poppy}, being rewritten from the ground up taking full advantage of more-recent \jax features and libraries.  It is built in an object-oriented \jax framework called \equinox \cite{kidger2021equinox}, with an optical system consisting of an \equinox module wrapping a stack of `layers' which are themselves \equinox modules applying phase or amplitude screens, or performing Fourier or Fresnel transforms on a wavefront. We use the optimization library \optax \cite{optax2020github} for gradient-descent optimization, and the probabilistic programming library \numpyro \cite{Phan2019} for \ac{hmc}.

% Autodiff Optics
There are several other frameworks for automatically-differentiable optics simulation:
% WaveBlocks
the most similar is WaveBlocks \cite{page2020}, a PyTorch package for object-oriented modelling of fluorescence microscopy. 
% Sitzmann etc
The DeepOptics package \cite{sitzmann2018} has been developed in TensorFlow for camera design and computational imaging, and
% Liaudat
WaveDiff in TensorFlow has been used for detector calibration \cite{liaudat2021} and \ac{psf} modelling \cite{Liaudat2022}. 
We believe that \dlux fills a unique niche in this ecosystem as the most general modelling framework for physical optics. This has driven a design focus on flexibility and ease of open-source contribution and development. The goal is to provide a general object-oriented framework that enables a new user to harness the advantages of autodiff and hardware acceleration provided by \jax with its user-friendly \textsc{NumPy}-like API.

@@

The optimisation is performed for a total of 100 epochs, with each value-and-gradient calculation taking $\sim$ 2.75 seconds to evaluate on an Apple~M1~Max CPU, for a total optimisation time of $\sim$ 4.5 minutes.

This optimisation strategy performs well, with all of the parameters being recovered down to the noise level. This can be seen in Figure~\ref{fig:data_residual} with the right panel showing a sample residual of the data and model after optimisation, which is clearly only noise. Correlations and residual plots of the positions and fluxes after the optimisation are shown in Figure~\ref{fig:astro_params}. The 1-$\sigma$ error bars 
are generated by autodiff, calculating the covariance matrix under the Laplace approximation as the negative matrix inverse of the Hessian of the likelihood function evaluated at the maximum likelihood estimate. These errorbars are clearly an underestimate: this is understandable as they are marginalised over positions, fluxes and aberration, but not the \ac{prf} residuals which are the dominant noise source. The million-dimensional Hessian required to estimate these uncertainties is at present intractable even in \jax. These parameters are recovered to the expected distribution when performed without modelling the \ac{prf}. Figure~\ref{fig:aberrations} shows the recovery of the optical aberrations, with a final residual aberration of 
% \textcolor{red}{showyourwork variable}
\variable{output/rms_opd_resid.txt} nm RMS. 
Finally, Figure~\ref{fig:flat_field} shows a correlation plot of the true and recovered sensitivity values and the corresponding residual histogram. The points of the correlation plot are colour-coded with the total counts for each pixel, showing as expected that pixels with higher photon counts having a better recovery of the PRF since these pixels have a greater \ac{snr}. Overall the parameters are very well recovered down to the noise level.

@@
